\documentclass{article}
\usepackage{hevea}

\def\t#1{{\tt #1}}
\def\DYNAMIC{\t{DYNAMIC}}
\title{Setting Up CVS, SSH and OCaml}
\author{Scott McPeak \and George Necula}

\begin{document}
\maketitle

 This document is intended to get you started with OCAML and the other tools
that are necessary for CCured and other projects of ours. These tools work on
Linux and Windows (NT4.0, 2000, XP and also less reliably on 95/98/Me).

 If you use Linux then you can go directly to Section~\ref{sec-ocaml}.

\section{If you want to use Windows}

 \subsection{Get \t{cygwin}}

 You must have a bunch of Unix tools installed on your machine. (In the future
we might be able to avoid these but for now you are better off with them.).
Here is what I (George) do to install Cygwin. You need a good network
connection for this. 
\begin{itemize}
\item Create a directory \t{C:\backslash Download\backslash cygwin}
\item Go to \ahrefurl{http://sources.redhat.com/cygwin} and click \t{Install
cygwin} icon. Download \t{setup.exe} to the directory you just created.
\item Run \t{setup.exe} and select ``Download to local directory''. Select all
the packages. This will take a while (~ 30 minutes)
\item Run \t{setup.exe} again and not select to ``Install from local
directory''. It is best to {\bf deselect} the \t{tetex} package since I found
it to interfere with other installations of Latex.
\item I choose \t{C:\backslash Programs\backslash cygwin} 
as the home for \t{cygwin}, I use \t{DOS} as the default text file and I
choose ``Install for All''. 
\item Add \t{C:\backslash Programs\backslash cygwin\backslash bin} to your
PATH. You must put it first in the ``System Variables'' PATH (In Control Panel/System/Advanced/Environment
Variables'' and {\bf put it first} so that it comes before the default
c:/WinNT/system32. You can verify that you got it right if you get
\t{/usr/bin/find} when you run \t{which find}. 
\item If you are on a Win NT/2k/XP you might want to set "CYGWIN=ntsec" so
that cygwin knows how to work with NTFS's file permissions.
\end{itemize}


\section{Get OCaml}\label{sec-ocaml}

The next step for most people is to download and install the Ocaml
compiler system.  This is available at:

  \ahrefurl{http://caml.inria.fr/ocaml/}

At the time of writing, the current version is 3.02. The following
instructions assume that you will build from sources even on Windows (instead
of using the Native compiler provided). Here the rough sequence of steps:
\begin{itemize}
\item Download and unpack the sources
\item Go in that directory, start \t{bash} if on Windows, and run
\begin{verbatim}
% ./configure
% make world
% make opt
% make install
\end{verbatim}

 The first command might fail to configure some libraries but that's Ok. 
\end{itemize}

To test your ocaml distribution, try:

\begin{verbatim}
  % which ocaml
  /usr/local/bin/ocaml

  % ocaml
        Objective Caml version 3.02

  # exit 0;;    <-- you type "exit 0;;", and press enter
\end{verbatim}


\section{Set the environment variables}

 Make sure that /usr/local/lib in in your path. If you are on Windows this
 means ``C:/Programs/cygwin/usr/local/lib''. 

 Set the \t{ARCHOS} variable to either \t{x86\_WIN32} (if you are on Windows)
 or to \t{x86\_LINUX} (if you are on Linux)


\section{If you need CVS access} 

 See the \ahref{cvssetup.html}{separate instructions}.


\section{Useful Links}
\begin{itemize}
 \item  Tutorial on ML: 
   \ahrefurl{http://www.dcs.napier.ac.uk/course-notes/sml/manual.html}

        
  \item Documentation and sources for Ocaml: 
        \ahrefurl{http://caml.inria.fr/ocaml/} 
  \item Manual for GNU make:
        \ahrefurl{http://www.gnu.org/manual/make/html\_chapter/make\_toc.html}

    
 \end{itemize}


\end{document}
